---
title: "BIOL 6220 Notebook"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

# R Basic
## Introduction

R as a calculator

```{r}
1 + 100
```
```{r}
2^2 
```

```{r}
7/2
```

```{r}
4-1 
```
```{r}
2*2
```
```{r}
3+5*2
```

```{r}
2e2
```

```{r}
sin(1)
log(1)
exp(0)

```

```{r}
1 == 1
```
```{r}
1 < 1 
```
```{r}
1 != 2
```
```{r}
x <- 10
x
```
### Vectorization

```{r}
1:5
```

```{r}
2^(1:5)
```

```{r}
v <- 1:5
2^v
```
### Environment


```{r}
ls()
``
```

## Package Management 




```{r}
installed.packages()
```

```{r}
install.packages("vegan", dependencies=TRUE)
```

```{r}
library(vegan)
remove.packages(vegan)











```

## Project Management 

1. Set up Management structure (i.e, "data","src","results","doc")
2. Discuss avoiding redundancy of files 
3. Make sure to use version control (i.e, Git)

## Getting Help
```{r}
?plot
```

```{r}
?"<-"
```


```{r}
vignette("FAQ-vegan")
```

```{r}
citation("vegan")
```

## Data and their formats 

```{r}
cats <- data.frame(coat=c("calico","black","tabby")), weight = c(2,1,5,3,2) likes_string = c(1,0,1)
                  
```

```{r}
cats
```

```{r}
write.csv(cats,file= "./data/cats.csv")
```


```{r}
cats$coat
``
```

```{r}
cbind(cats,logweight)
```










```{r}
typeof(FALSE)
```
```{r}
typeof(T)
typeof(F)
```
```{r}
class(cats)
```
```{r}
file.show("./data/cats.csv")
```

```{r}
c("a","b")
```

```{r}
ab <- c("a","b")

```

```{r}
ab
```

```{r}
1:10
```
```{r}
z <- seq(10)
```

```{r}
seq(1,100,by=5)
```



### Data Frames

```{r}
cats
```



```{r}
str(cats$coat)
```

```{r}
coats <- c("tabby","tortoiseshell","tortoiseshell","black","tabby")

```

```{r}
coats
```



```{r}
factor(coats)
```

```{r}
categories <- factor(coats)
class(coats)
class(categories)
```

```{r}
str(categories)
```

### Lists

```{r}
list_example <- list(title="numbers", numbers = 1:10, data=TRUE)
```

```{r}
list_example
```
```{r}
another_list <- list(1,"a",TRUE,1+1i)
another_list
```

```{r}
typeof(list_example)
```
```{r}
data.frame(list_example)
```

```{r}

```
































### Matrices 


```{r}
matrix_example <- matrix(0, ncol=5, nrow=3)
```

```{r}
matrix_example
```

```{r}
class(matrix_example)
```

```{r}
typeof(matrix_example)
```

```{r}
str(matrix_example)
```

```{r}
dim(matrix_example)
```

```{r}
ncol(matrix_example)
```

```{r}
nrow(matrix_example)
```

```{r}
class(data.frame(matrix_example))
```

```{r}
df_example <- data.frame(matrix_example)
```
```{r}
df_example
```

### subsetting 

```{r}
p <- c(2.3,6.9,4,0,23,1)
```

```{r}
names(p) <- c('a','b','c','d','e')
p
```

```{r}
p[1]
```
```{r}
p[1:3]
```

```{r}
p[c(1,5)]
```

```{r}
p[c(1,1,1,3,5,5)]
```

```{r}
p[-3]
```

```{r}
p[c(-1,-5)]
```

```{r}
p[c('a','c')]
```
```{r}
p[names(p) != 'c']
```

### Factors

```{r}
f <- factor(c('a','b','c','d','e'))
f
```

```{r}
f [f == 'a']
```

```{r}
f[1:3]
```

```{r}
f[f %in% c('b','c')]
```

```{r}
f[-3]
```

```{r}
f2 <- factor(c('a','a','b','c','c'))
f2
```

```{r}
f2[f2 == 'a']
```

```{r}
f2[f2 %in% c('a','c')]
```







### Matrices Resumed

```{r}
set.seed(65)
m <- matrix(rnorm(6*4), ncol=4, nrow=6)
m
```

```{r}
m[3:4, c(3,1)]
```

```{r}
m[,c(3,4)]
```
###Lists Again
```{r}
xlist <- list(a="BIOL6220",b=seq(1,10,by = 0.5), data="Grade")
xlist
```

```{r}
xlist[1:2]
```

```{r}
xlist['a']
```

```{r}
xlist[['a']]
```

```{r}
xlist['b']
```
```{r}
xlist[['b']]
```

```{r}
xlist$b
```

### Data Frames Revisited 


```{r}
gp <- read.csv('./data/forest_area_sq_km.csv')
```



```{r}
head(gp, n=10L)
```


```{r}
head(gp[3],n=10)
```


```{r}
nrow(gp)

```


```{r}
head(gp[["country"]], 10L)
```

```{r}
gp$X2001 
```


```{r}
gp[1:3, 2:5]
```

```{r}
gp[ which(gp$X1989 <= 100 & gp$X2019 >= 100),]
```

### Conditionals and Flow 









```{r}
n <- 10

if(n < 10) { 
  print("n is less than 10")
  
} else if (n >10) { 
    print("n is greater than 10")
}else { 
    "n is equal to 10"
  }
```

```{r}


for (n in seq(0,20)) {
  if(n < 10) { 
    print("n is less than 10")
  
  } else if (n >10) { 
      print("n is greater than 10")
  }else { 
      "n is equal to 10"
  }
}  
```

```{r}
g <- 0

while (g<=10) {
  print(paste(g, "is less than or equal to 10"))
  g <- g+1 
}
```

## Plotting and Figures

```{r}
#install.packages("ggplot2")
```


```{r}
library(ggplot2)
```




```{r}
gp <- read.csv('data/gapminder_all.csv')

head(gp)
```

```{r}
ggplot(data=gp, mapping=aes(x=gdpPercap_1952, y=pop_1952)) + geom_point()
```



```{r}
ggplot(data=gp, mapping=aes(x=gdpPercap_2002, y=pop_2002)) + geom_point()
```



```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv")
head(gapminder)
```


```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp))+ geom_point()
```


```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country, color=continent))+ geom_line()
```



```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country)) + geom_line(mapping=aes(color=continent)) +geom_point()
```





```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp))+ geom_point(alpha=0.25) + scale_x_log10()
```


```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp))+ geom_point(alpha=0.25, color="purple", size = 0.9, pch = 9)+ scale_x_log10() + geom_smooth(method = lm, color="gold", size= 1.25)
```


```{r}
africas <- gapminder[gapminder$continent == "Africa",]
head(africas)
```


```{r}
ggplot(data=africas, mapping=aes(x=year, y=lifeExp)) + 
  geom_line()+
  facet_wrap(~ country)+
  theme(axis.text.x = element_text(angle=90)) + 
  labs(
    x= "Year",
    y = "Life Expectency",
    title = "Life Expectency Over Time in African Countries",
    color = "red"
  )
```



```{r}
AfricanLifeExp <- ggplot(data=africas, mapping=aes(x=year, y=lifeExp)) + 
  geom_line()+
  facet_wrap(~ country)+
  theme(axis.text.x = element_text(angle=90)) + 
  labs(
    x= "Year",
    y = "Life Expectency",
    title = "Life Expectency Over Time in African Countries",
    color = "red"
  )
```

```{r}
ggsave(filename= "data/AfricanLifeExp.png", plot = AfricanLifeExp, width = 24, height = 40, dpi = 300, units = "cm")
```


```{r}
pdf(file = "results/AfricanLifeExp.pdf", width = 24, height = 40)
plot(AfricanLifeExp)
dev.off()
```

```{r}
write.table(gapminder, file="data/gapminder_web.csv", sep=",")
```

```{r}
write.csv(africas, file="data/gapminder_web_africas.csv")`
```

